---
title: "Water Physics"
author: "Jacob J. Boulrice"
date: "2025-03-23"
output: html_document
---

```{r setup, warning=FALSE, message=FALSE}
# libraries
library(readr)
library(ggplot2)
library(dplyr)
```

	In the following experiment, participants are tasked with launching a ball across a water current in a virtual environment (VE), with the goal of hitting a target. To succeed, participants must learn the physics parameters of the VE and how these forces interact with the ball’s trajectory, to refine their launch parameters (angle and speed). 
	
## Load Simulation Data:
```{r}
launch_data_n2 <- read_csv("C:/Users/jacob/OneDrive/Desktop/SMC LAB/Projects/Water-Current-Analysis/simulation_data/launch_data_-2.0/launch_data_-2.0.csv", show_col_types = FALSE)

```
## Organize Simulation Data:
```{r}
# Add factor water_speed_m_s to each df
launch_data_n2$water_speed_m_s <- factor(rep('-2',nrow(launch_data_n2)))

# check
head(launch_data_n2)
```


```{r}
# Combine individual water speed dfs
launch_data_total <- rbind(launch_data_n2)


# round summary df to 2 decimal points
launch_data_total <- launch_data_total %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

```


```{r}
# Change target_angle, target_x_pos, target_z_pos, and target_radius to factors
launch_data_total <- launch_data_total %>% mutate(across(c(
  target_angle, target_x_pos, target_z_pos, target_radius), as.factor))

# check
head(launch_data_total)
```
## Summarize Simulation Data and Create New Dataframe:
```{r}
summary_launch_data <- launch_data_total %>%
  group_by(target_angle, target_x_pos, target_z_pos, target_radius, water_speed_m_s) %>%
    # hit rate
    mutate(hit_rate = mean(target_hit)*100) %>%
  filter(target_hit == 'TRUE') %>%
  group_by(target_angle, target_x_pos, target_z_pos, target_radius, water_speed_m_s, hit_rate) %>%
  summarise(
    # minimum speed to hit the target
    min_speed_hit = min(launch_speed_m_s),
    # span of angles that hit the target
    launch_angle_range_hit = max(launch_angle_deg) - min(launch_angle_deg),
    # launch angle at minimum speed
    launch_angle_at_min_speed = launch_angle_deg[which.min(launch_speed_m_s)],
    # Difference between target angle and launch angle at minimum speed
    min_launch_angle_delta = ( as.numeric(as.character(first(target_angle))) - launch_angle_at_min_speed),
    # total number of target hits for each parameter
    total_hits = n(),
  ) %>%
  ungroup()

# check
head(summary_launch_data)

# export df to csv
write_csv(summary_launch_data, file = 'summary_launch_data.csv')
```

## Water Speed (m/s) = -1.5
```{r}
summary_launch_data_n2 <- summary_launch_data %>%
  filter(water_speed_m_s == -2) %>%
  filter(target_z_pos == 1.4) %>%
  group_by(target_angle, target_x_pos, target_z_pos, target_radius)


write_csv(summary_launch_data_n2, "summary_sim_data.csv")
```


# Total Hits Figures
```{r}
 #target_x_labs <- c("-0.5", "-0.25", "0")
 #target_z_labs <- c("-0.5", "-0.25", "0")

# Total hits vs Target radius 
ggplot(summary_launch_data_n2, 
       aes(x = target_radius, y = total_hits, 
           shape = target_z_pos, 
           color = target_x_pos, 
           fill = target_x_pos)) +
  geom_jitter(width = 0, size = 3, alpha = 0.75) +
  theme_minimal() +
  labs(title = "Total Hits vs Target Radius", x = "Target Radius", y = "Total Hits")


# Hit Rate vs Target radius 
ggplot(summary_launch_data_n2, 
       aes(x = target_radius, y = hit_rate, 
           shape = target_z_pos, 
           color = target_x_pos, 
           fill = target_x_pos)) +
  geom_jitter(width = 0, size = 3, alpha = 0.75) +
  theme_minimal() +
  labs(title = "Hit Rate vs Target Radius", x = "Target Radius", y = "Hit Rate")
```

# Minimum Speed Figures
```{r}
# Min Speed vs Target radius 
ggplot(summary_launch_data_n2, 
       aes(x = target_radius, y = min_speed_hit, 
           shape = target_z_pos, 
           color = target_x_pos, 
           fill = target_x_pos)) +
  geom_jitter(width = 0, size = 3, alpha = 0.75) +
  theme_minimal() +
  labs(title = "Min Speed vs Target Radius", x = "Target Radius", y = "Min Speed")
```

# Launch Angle Range Figures
```{r}
# Launch Angle Span vs Target radius 
ggplot(summary_launch_data_n2, 
       aes(x = target_radius, y = launch_angle_range_hit, 
           shape = target_z_pos, 
           color = target_x_pos, 
           fill = target_x_pos)) +
  geom_jitter(width = 0, size = 3, alpha = 0.75) +
  facet_wrap(~ water_speed_m_s) +
  theme_minimal() +
  labs(title = "Launch Angle Range vs Target Radius", x = "Target Radius", y = "Launch Angle Range")
```

# Launch Angle at Minimum Speed Figures
```{r}
ggplot(summary_launch_data_n2, 
       aes(x = target_radius, y = launch_angle_at_min_speed, 
           shape = target_z_pos, 
           color = target_x_pos, 
           fill = target_x_pos)) +
  geom_jitter(width = 0, size = 3, alpha = 0.75) +
  facet_wrap(~ water_speed_m_s) +
  theme_minimal() +
  labs(title = "Launch Angle at Min Speed vs Target Radius", x = "Target Radius", y = "Launch Angle at Min Speed")
```

# Angular Difference from Minimum Speed Figures
```{r}
ggplot(summary_launch_data_n2, 
       aes(x = target_radius, y = min_launch_angle_delta, 
           shape = target_z_pos, 
           color = target_x_pos, 
           fill = target_x_pos)) +
  geom_jitter(width = 0, size = 3, alpha = 0.75) +
  facet_wrap(~ water_speed_m_s) +
  theme_minimal() +
  labs(title = "Angular difference at Min Speed vs Target Radius", x = "Target Radius", y = "Angular difference at Min Speed")
```



## Participant Data
```{r}
# load participant data
dat_1 <- read_csv("Data_Summer_25/trial_results_jacob_b.csv")

# Combine data sets
combined_df <- bind_rows(dat_1, dat_2, dat_3, dat_4, dat_5, dat_6, dat_7, dat_8, dat_9)

combined_df <- combined_df %>%
  mutate(target_id = case_when(
    target_position_x == 0   & target_position_z == 1.5  & target_width == 0.151131 ~ "big_target",
    target_position_x == 0   & target_position_z == 1.25 & target_width == 0.100754 ~ "medium_target",
    target_position_x == -0.5 & target_position_z == 1.4  & target_width == 0.075566 ~ "small_target",
    TRUE ~ "unknown"
  ))

combined_df$target_id <- factor(combined_df$target_id)
str(combined_df$target_id)

combined_df$ppid <- factor(combined_df$ppid)
levels(combined_df$ppid)

combined_df$water_speed_binary <- ifelse(combined_df$water_speed_m_s == -2, 1, 0)
combined_df$water_speed_binary <- factor(combined_df$water_speed_binary)
str(combined_df$water_speed_binary)

combined_df$target_angle_binary <- ifelse(combined_df$target_angle == -9.7, 1, 0)



combined_df <- combined_df %>%
  mutate(width_order_group = case_when(
    ppid %in% c("p1_t1", "p3_t1", "p5_t1", "p7_t1", "p9_t1") ~ "Wide to Narrow",
    ppid %in% c("p1_t3", "p4_t3", "p6_t3", "p7_t3") ~ "Narrow to Wide",
    TRUE ~ "Other"
  ))

combined_df$width_order_group <- factor(combined_df$width_order_group)
str(combined_df$width_order_group)

unique(combined_df$width_order_group)

combined_df <- combined_df %>%
  group_by(ppid) %>%
  mutate(starting_offcentered = first(target_angle_binary)) %>%
  ungroup()

combined_df <- combined_df %>%
  group_by(target_angle) %>%
  mutate(target_angle_mt = abs(target_angle) + 90) %>%
  ungroup()

combined_df <- combined_df %>%
  group_by(ppid) %>%
  mutate(target_angle_dev = launch_angle - target_angle_mt) %>%
  ungroup()
# where target angle dev tells us how much an individual trial launch angle deviates from the true target angle, relative to the origin.
# Where a launch angle of 90 degrees is directly in front of the participant


combined_df <- combined_df %>%
  mutate(
    # X is already relative; Z needs centering
    distance_to_center = sqrt((min_pos_from_target_x)^2 + 
                              (min_pos_from_target_z - target_position_z)^2),
    normalized_distance = distance_to_center - target_width
  )

# scale distance
combined_df$distance_to_center_std <- scale(combined_df$distance_to_center)

combined_df_water <- combined_df %>%
  filter(water_speed_m_s != 0)

```



```{r}
combined_df_ft <- combined_df %>%
  arrange(ppid, trial_num) %>%  # Ensure trials are ordered
  group_by(ppid) %>%
  mutate(
    prev_water = lag(water_speed_binary, default = first(water_speed_binary)),
    transition_to_1 = water_speed_binary == 1 & prev_water == 0,
    to_delete = transition_to_1 & row_number() > 1  # avoid first row corner case
  ) %>%
  ungroup() %>%
  filter(!to_delete)  # Remove those transition rows

# Tag the rows to delete before filtering
deleted_trials <- combined_df_ft %>%
  arrange(ppid, trial_num) %>%
  group_by(ppid) %>%
  mutate(
    prev_water = lag(water_speed_binary, default = first(water_speed_binary)),
    transition_to_1 = water_speed_binary == 1 & prev_water == 0,
    to_delete = transition_to_1 & row_number() > 1
  ) %>%
  ungroup() %>%
  filter(to_delete) %>%
  select(ppid, trial_num, water_speed_binary, prev_water)

# Print trial numbers 
print(deleted_trials$trial_num)


combined_df_ft_water <- combined_df_water %>%
  filter(water_speed_m_s != 0)
```

```{r}
ggplot(combined_df_ft, aes(x = trial_num, y = distance_to_center)) +
  geom_line() +
  facet_wrap(~width_order_group)

ggplot(combined_df_ft, aes(x = trial_num, y = distance_to_center)) +
  geom_line() +
  facet_wrap(~ppid)

```


```{r}
ggplot(combined_df_ft_water, aes(x = trial_num, y = normalized_distance, color = target_id)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~width_order_group) +
  labs(title = "Normalized Distance Over Trials by Target Size",
       x = "Trial Number", y = "Normalized Distance") +
  theme_minimal()

combined_df_water %>%
  group_by(trial_num, target_id, width_order_group) %>%
  summarize(mean_dist = mean(distance_to_center, na.rm = TRUE)) %>%
  ggplot(aes(x = trial_num, y = mean_dist, color = target_id)) +
  
  geom_line(size = 1, alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, span = 0.4) +
  facet_grid(~width_order_group) +
  labs(
    title = "Mean Distance to Target Across Trials",
    x = "Trial Number",
    y = "Mean Distance (m)",
    color = "Target Condition"
  ) +
  theme_minimal()


```
## LOESS
```{r}
ggplot(combined_df, aes(x = trial_num, y = launch_angle)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.3, se = TRUE, color = "blue") +
  facet_wrap(~ width_order_group) +
  labs(
    x = "Trial Number",
    y = "Launch Angle",
    title = "Launch Angle vs. Trial Number by Width Order Group (LOESS)"
  ) +
  theme_minimal()
ggsave("loess_angle_trials.png", dpi = 300, width = 12, height = 8, bg = 'white')


ggplot(combined_df, aes(x = trial_num, y = launch_Speed)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.3, se = TRUE, color = "blue") +
  facet_wrap(~ width_order_group) +
  labs(
    x = "Trial Number",
    y = "Launch Speed",
    title = "Launch Speed vs. Trial Number by Width Order Group (LOESS)"
  ) +
  theme_minimal()
ggsave("loess_speed_trials.png", dpi = 300, width = 12, height = 8, bg = 'white')
```


```{r}
ggplot(combined_df_water, aes(x = block_num, y = normalized_distance, color = target_id)) +
  geom_point(alpha = 0.6) +  # To show individual points for each trial
  geom_line(aes(group = target_id), alpha = 0.8) +  # Optional: Connect trials within each target condition
  facet_wrap(~ width_order_group) +  # Separate by target condition (big, medium, small)
  labs(
    title = "Normalized Distance to Target Over Trials (with current)",
    x = "Trial Number",
    y = "Distance to Target (m)",
    color = "Target Condition"
  ) +
  theme_minimal() +
  theme(legend.position = "top")


ggsave("min_euclidian_dist_by_group.png", dpi = 300, width = 12, height = 8, bg = 'white')

```



# Linear Model: Speed
```{r}

mod_speed <- lm(launch_Speed ~ water_speed_binary * trial_num + width_order_group + target_width, data = combined_df)
summary(mod_speed)
```


# mixed model: speed
```{r}

library(lme4)
mx_mod_speed <- lmer(launch_Speed ~ water_speed_binary * block_num + launch_angle + target_position_x + target_position_z + starting_offcentered * water_speed_binary + (1 + block_num|ppid), data = combined_df, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))



summary(mx_mod_speed)

#AIC(mx_mod_speed, mod_speed)
#BIC(mx_mod_speed, mod_speed)
```


```{r}
# Open PNG
png("mixed_model_diagnostics_speed.png", width = 1600, height = 1200, res = 200)

# 2x2 layout for plots
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Residuals vs Fitted
plot(fitted(mx_mod_speed), residuals(mx_mod_speed),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 20, col = "#1f77b4")
abline(h = 0, col = "gray")

# QQ Plot
qqnorm(residuals(mx_mod_speed), main = "Normal Q-Q",
       pch = 20, col = "#ff7f0e")
qqline(residuals(mx_mod_speed), col = "gray")

# Scale-Location Plot
plot(fitted(mx_mod_angle_block_speed),
     sqrt(abs(residuals(mx_mod_speed))),
     main = "Scale-Location",
     xlab = "Fitted Values",
     ylab = "Sqrt(|Residuals|)",
     pch = 20, col = "#2ca02c")
abline(h = 0, col = "gray")

# Histogram of Residuals
hist(residuals(mx_mod_speed),
     breaks = 50,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "skyblue", border = "white")

# Close and save file
dev.off()
```
```{r}
library(performance)
r2(mx_mod_speed)

confint(mx_mod_speed, method = "boot", nsim = 1000)
```
```{r}

# Create prediction grid
other_vars <- expand.grid(
  block_num = seq(1, 60, by = 1),
  water_speed_binary = factor(c(0, 1), levels = c(0,1)),
  launch_angle = mean(combined_df$launch_angle),
  ppid = unique(combined_df$ppid),
  #starting_offcentered = unique(combined_df$starting_offcentered)
  starting_offcentered = 0
)

target_combos <- data.frame(
  target_position_x = c(0, 0, -0.5),
  target_position_z = c(1.5, 1.25, 1.4)
)

newdata <- merge(other_vars, target_combos)

# Get predicted launch angles from model with random ppid effects
newdata$pred_speed <- predict(mx_mod_speed, newdata = newdata)

# label water condition
newdata$water_condition <- ifelse(newdata$water_speed_binary == 1, "Water Current", "Still-Water")

newdata <- newdata %>%
  mutate(target_label = paste0("x=", target_position_x, ", z=", target_position_z),
         target_label = factor(target_label))  # plotting labels

# Plot
ggplot(newdata, aes(x = block_num, y = pred_speed,
                    color = target_label, linetype = water_condition,
             group = interaction(ppid, target_label, water_condition))) +
  geom_line(size = 1) +
  facet_wrap(~ ppid, ncol = 3) +
  labs(title = "Predicted Launch Speed Over Blocks",
       subtitle = "From Mixed Model: Speed",
       x = "Block Number",
       y = "Predicted Launch Speed (m/s)",
       color = "Target (x,z)",
       linetype = "Water Condition") +
  theme_minimal(base_size = 12)

ggsave("predicted_launch_speed_by_participant.png", dpi = 300, width = 12, height = 8, bg = 'white')



# Plot
ggplot(newdata, aes(x = block_num, y = pred_speed,
                         color = water_speed_binary, group = water_speed_binary)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2, aes(fill = water_speed_binary)) +
  scale_color_manual(values = c("red", "blue"), labels = c("Still Water", "Water Current")) + 
  scale_fill_manual(values = c("red", "blue"), labels = c("Still Water", "Water Current")) +
  labs(title = "Group-Level Launch Speed Over Blocks",
       x = "Block Number", y = "Predicted Launch Speed (m/s)",
       color = "Water Condition", fill = "Water Condition") +
  theme_minimal()

ggsave("predicted_launch_speed_by_water_current.png", dpi = 300, width = 12, height = 8, bg = 'white')


```


# Linear Model: Angle
```{r}
mod_angle <- lm(launch_angle ~ water_speed_binary * trial_num + width_order_group + target_angle_binary, data = combined_df)
summary(mod_angle)

plot(mod_angle)

mod_test <- lm(launch_angle ~ water_speed_binary, data = combined_df)
summary(mod_test)

# Add predicted values for mod_angle
combined_df$predicted_test <- predict(mod_test, newdata = combined_df)
str(combined_df$predicted_test)

ggplot(combined_df, aes(x = predicted_test, y = launch_angle)) +
  geom_point(alpha = 0.5) +
  facet_grid(~water_speed_binary) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Predicted vs. Observed Launch Angle",
       x = "Predicted Launch Angle",
       y = "Observed Launch Angle") +
  theme_minimal()

combined_df$residuals <- resid(mod_test)

ggplot(combined_df, aes(x = as.factor(water_speed_binary), y = residuals)) +
  geom_boxplot() +
  labs(x = "Water Speed Condition", y = "Residuals") +
  theme_minimal()

car::leveneTest(residuals ~ factor(water_speed_binary), data = combined_df)

```

```{r}

library(car)
vif(mod_angle, type = 'predictor')

# Extract residuals
resid_mod_angle <- residuals(mod_angle)

# testing correlation with residuals (model mispecification)
cor.test(combined_df$trial_num, resid_mod_angle)
cor.test(combined_df$water_speed_binary, resid_mod_angle)
cor.test(combined_df$target_angle_binary, resid_mod_angle)



# Add predicted values for mod_angle
combined_df$predicted_angle <- predict(mod_angle, newdata = combined_df)
str(combined_df$predicted_angle)

ggplot(combined_df, aes(x = predicted_angle, y = launch_angle)) +
  geom_point(alpha = 0.5) +
  facet_grid(~water_speed_binary) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Predicted vs. Observed Launch Angle",
       x = "Predicted Launch Angle",
       y = "Observed Launch Angle") +
  theme_minimal()

```
# Standardize key variables
```{r}
combined_df$target_position_z_std <- scale(combined_df$target_position_z)
unique(combined_df$target_position_z_std)
mean(combined_df$target_position_z)
```


## Mixed Models: Angle
```{r}
library(lme4)
library(ggeffects)

# mixed model
mx_mod_angle_block_speed <- lmer(launch_angle ~ water_speed_binary * block_num + launch_Speed + target_position_x + target_position_z + starting_offcentered * water_speed_binary + (1 + block_num|ppid), data = combined_df, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
summary(mx_mod_angle_block_speed)

# comparing mixed model to original model
#AIC(mx_mod_angle_block_speed, mod_test)
#BIC(mx_mod_angle_block_speed, mod_test)



# mixed model
mx_mod_angle_test <- lmer(launch_angle ~ water_speed_binary * block_num + launch_Speed + target_position_x + target_position_z + starting_offcentered * water_speed_binary * launch_Speed + (1 + block_num|ppid), data = combined_df, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
summary(mx_mod_angle_test)

# comparing mixed model to original model
AIC(mx_mod_angle_block_speed, mx_mod_angle_test)
BIC(mx_mod_angle_block_speed, mx_mod_angle_test)

library(performance)
r2(mx_mod_angle_block_speed)
r2(mx_mod_angle_test)
```



```{r}
# Open PNG
png("mixed_model_diagnostics.png", width = 1600, height = 1200, res = 200)

# 2x2 layout for plots
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Residuals vs Fitted
plot(fitted(mx_mod_angle_block_speed), residuals(mx_mod_angle_block_speed),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 20, col = "#1f77b4")
abline(h = 0, col = "gray")

# QQ Plot
qqnorm(residuals(mx_mod_angle_block_speed), main = "Normal Q-Q",
       pch = 20, col = "#ff7f0e")
qqline(residuals(mx_mod_angle_block_speed), col = "gray")

# Scale-Location Plot
plot(fitted(mx_mod_angle_block_speed),
     sqrt(abs(residuals(mx_mod_angle_block_speed))),
     main = "Scale-Location",
     xlab = "Fitted Values",
     ylab = "Sqrt(|Residuals|)",
     pch = 20, col = "#2ca02c")
abline(h = 0, col = "gray")

# Histogram of Residuals
hist(residuals(mx_mod_angle_block_speed),
     breaks = 50,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "skyblue", border = "white")

# Close and save file
dev.off()
```
## Launch Angle Mixed Effects Model Equation:
$$
\begin{aligned}
\hat{y}_{ij} \;=\;& 99.02221 \;-\; 32.34682\,\bigl(\text{WaterSpeedBinary}_{ij}\bigr)
                \;-\; 0.12076\,\bigl(\text{BlockNumber}_{ij}\bigr) \\
                &\;+\; 5.07287\,\bigl(\text{LaunchSpeed}_{ij}\bigr)
                \;-\; 36.00173\,\bigl(\text{TargetX}_{ij}\bigr) \\
                &\;-\; 11.96141\,\bigl(\text{TargetZ}_{ij}\bigr)
                \;-\; 0.22356\,\bigl(\text{StartingOffcentered}_{ij}\bigr) \\
                &\;+\; 0.23348\,\bigl(\text{WaterSpeedBinary}_{ij} \times \text{BlockNumber}_{ij}\bigr) \\
                &\;-\; 5.74310\,\bigl(\text{WaterSpeedBinary}_{ij} \times \text{StartingOffcentered}_{ij}\bigr) \\
                &\;+\; b_{0i} \;+\; b_{1i}\,\bigl(\text{BlockNumber}_{ij}\bigr) \;+\; \epsilon_{ij}.
\end{aligned}
$$

```{r}
r2(mx_mod_angle_block_speed)

confint(mx_mod_angle_block_speed, method = "boot", nsim = 1000)
```

```{r}

# Create prediction grid
other_vars <- expand.grid(
  block_num = seq(1, 60, by = 1),
  water_speed_binary = factor(c(0, 1), levels = c(0,1)),
  launch_Speed = mean(combined_df$launch_Speed),
  ppid = unique(combined_df$ppid),
  starting_offcentered = unique(combined_df$starting_offcentered)
)

target_combos <- data.frame(
  target_position_x = c(0, 0, -0.5),
  target_position_z = c(1.5, 1.25, 1.4)
)

newdata <- merge(other_vars, target_combos)

# Get predicted launch angles from model with random ppid effects
newdata$pred_angle <- predict(mx_mod_angle_block_speed, newdata = newdata)

# label water condition
newdata$water_condition <- ifelse(newdata$water_speed_binary == 1, "Water Current", "Still-Water")

newdata <- newdata %>%
  mutate(target_label = paste0("x=", target_position_x, ", z=", target_position_z),
         target_label = factor(target_label))  # plotting labels

newdata <- newdata %>% 
  filter(
    (target_position_x == 0 & starting_offcentered == 0) |
      (target_position_x != 0 & starting_offcentered == 1)
  )

# Plot
ggplot(newdata, aes(x = block_num, y = pred_angle,
                    color = target_label, linetype = water_condition,
             group = interaction(ppid, target_label, water_condition))) +
  geom_line(size = 1) +
  facet_wrap(~ ppid, ncol = 3) +
  labs(title = "Predicted Launch Angle Over Blocks",
       subtitle = "From Mixed Model: Angle",
       x = "Block Number",
       y = "Predicted Launch Angle (degrees)",
       color = "Target (x,z)",
       linetype = "Water Condition") +
  theme_minimal(base_size = 12)

ggsave("predicted_launch_angles_by_participant.png", dpi = 300, width = 12, height = 8, bg = 'white')

# Plot
ggplot(newdata, aes(x = block_num, y = pred_angle,
                         color = water_speed_binary, group = water_speed_binary)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.2, aes(fill = water_speed_binary)) +
  scale_color_manual(values = c("red", "blue"), labels = c("Still Water", "Water Current")) + 
  scale_fill_manual(values = c("red", "blue"), labels = c("Still Water", "Water Current")) +
  labs(title = "Group-Level Launch Angle Over Blocks",
       x = "Block Number", y = "Predicted Launch Angle (°)",
       color = "Water Condition", fill = "Water Condition") +
  theme_minimal()

ggsave("predicted_launch_angles_by_water_current.png", dpi = 300, width = 12, height = 8, bg = 'white')

```
```{r}
# Create a prediction grid varying block and launch speed
grid <- expand.grid(
  block_num = seq(min(combined_df$block_num), max(combined_df$block_num), length.out = 60),
  launch_Speed = seq(min(combined_df$launch_Speed), max(combined_df$launch_Speed), length.out = 10),
  water_speed_binary = factor(c(0, 1), levels = c(0, 1))
)

# Set fixed values for other predictors
grid$target_position_x <- 0        # or another value of interest
grid$target_position_z <- 1.5        # or your chosen value
grid$starting_offcentered <- ifelse(grid$target_position_x == 0, 0, 1)
grid$ppid <- unique(combined_df$ppid)[1]  # use a reference participant

# Generate model predictions (using fixed effects only)
grid$pred_angle <- predict(mx_mod_angle_block_speed, newdata = grid, re.form = NA)

# Convert water condition for labeling
grid$water_condition <- ifelse(grid$water_speed_binary == 1, "Water Current", "Still Water")

# Plot using ggplot2
ggplot(grid, aes(x = block_num, y = pred_angle, color = as.factor(launch_Speed))) +
  geom_line(aes(group = launch_Speed), alpha = 0.7) +
  facet_wrap(~ water_condition, ncol = 1) +
  scale_color_viridis_d(name = "Launch Speed (binned)") +
  labs(title = "Predicted Launch Angle Across Blocks",
       subtitle = "Faceted by Water Condition; color indicates launch speed",
       x = "Block (Time)",
       y = "Predicted Launch Angle (degrees)") +
  theme_minimal(base_size = 12)

```





```{r}
mod_angle_shift <- lmer(target_angle_dev ~ water_speed_binary * block_num * width_order_group * launch_Speed + target_position_x + target_position_z + (1 + block_num|ppid), data = combined_df, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
summary(mod_angle_shift)

mod_angle_shift_test_1 <- lmer(target_angle_dev ~ water_speed_binary * block_num * width_order_group * launch_Speed + target_position_x + target_position_z + (1 + block_num|ppid), data = combined_df, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
summary(mod_angle_shift)

AIC(mod_angle_shift, mod_angle_shift_test_1)
BIC(mod_angle_shift, mod_angle_shift_test_1)


r2(mod_angle_shift)
r2(mod_angle_shift_test_1)

ggplot(combined_df, aes(x = trial_num, y = target_angle_dev)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.3, se = TRUE, color = "blue") +
  facet_wrap(~ width_order_group) +
  labs(
    x = "Trial Number",
    y = "Launch Angle-Target Deviation (degrees)",
    title = "Launch Angle-Target Deviation vs. Trial Number by Width Order Group (LOESS)"
  ) +
  theme_minimal()
ggsave("loess_deviation_trials.png", dpi = 300, width = 12, height = 8, bg = 'white')

```



```{r}
ggplot(combined_df, aes(x = launch_angle, y = launch_Speed, color = trial_num)) +
  geom_point(alpha = 0.5, size = 1.2) +
  scale_color_viridis_c(option = "plasma", name = "Trial #") +  # cool gradient
  facet_wrap(~width_order_group) +
  labs(
    title = "Trajectory Through 2D Solution Space",
    subtitle = "Angle vs Speed by Trial, Split by Group",
    x = "Launch Angle (degrees)",
    y = "Launch Speed (m/s)"
  ) +
  theme_minimal(base_size = 14)

ggsave("launch_angle_speed_by_trial_by_group.png", dpi = 300, width = 10, height = 6, bg = 'white')

```
# std key vars
```{r}
combined_df <- combined_df %>%
  mutate(
    block_num_z         = scale(block_num),
    launch_Speed_z      = scale(launch_Speed),
    target_angle_dev_z  = scale(target_angle_dev),
    launch_angle_z = scale(launch_angle),
    target_width        = factor(target_width),       
    width_order_group   = factor(width_order_group)     
  )


```


# Performance Model:

```{r}
mod_performance <- lmer(
  distance_to_center ~ water_speed_binary * block_num +
    poly(launch_Speed, 2) * water_speed_binary +
    poly(target_angle_dev, 2) * water_speed_binary +
    (1 | ppid),
  data = combined_df,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)


summary(mod_performance)

r2(mod_performance)

```


```{r}
# Open PNG
png("mixed_model_diagnostics_performance.png", width = 1600, height = 1200, res = 200)

# 2x2 layout for plots
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Residuals vs Fitted
plot(fitted(mod_performance), residuals(mod_performance),
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 20, col = "#1f77b4")
abline(h = 0, col = "gray")

# QQ Plot
qqnorm(residuals(mod_performance), main = "Normal Q-Q",
       pch = 20, col = "#ff7f0e")
qqline(residuals(mod_performance), col = "gray")

# Scale-Location Plot
plot(fitted(mod_performance),
     sqrt(abs(residuals(mod_performance))),
     main = "Scale-Location",
     xlab = "Fitted Values",
     ylab = "Sqrt(|Residuals|)",
     pch = 20, col = "#2ca02c")
abline(h = 0, col = "gray")

# Histogram of Residuals
hist(residuals(mod_performance),
     breaks = 50,
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "skyblue", border = "white")

# Close and save file
dev.off()
```

```{r}

library(ggeffects)   # for ggpredict()
library(ggplot2)     # for plotting

# Generate model predictions across target_angle_dev_z, separated by water condition
preds_angle <- ggpredict(mod_test, 
                   terms = c("target_angle_dev [all]", "water_speed_binary"))

preds_speed <- ggpredict(mod_test, 
                   terms = c("launch_Speed [all]", "water_speed_binary"))


# Plot
ggplot(preds_angle, aes(x = x, y = predicted, color = group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.2, color = NA) +
  labs(
    title = "Strategy Shift in Aiming (Angle Deviation) Across Water Conditions",
    x = "Target Angle Deviation (degrees)",
    y = "Predicted Distance to Center",
    color = "Water Current",
    fill = "Water Current"
  ) +
  theme_minimal(base_size = 14)

ggplot(preds_speed, aes(x = x, y = predicted, color = group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.2, color = NA) +
  labs(
    title = "Launch Speed Across Water Conditions",
    x = "Launch Speed (m/s)",
    y = "Predicted Distance to Center",
    color = "Water Current",
    fill = "Water Current"
  ) +
  theme_minimal(base_size = 14)

```

